\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{multirow}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10}
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{bn-en-translate: A Local GPU-Accelerated Bengali-to-English\\
Story Translation System with LoRA Fine-Tuning\\
and Real-Time Resource Monitoring}

\author{
  \IEEEauthorblockN{Souradeepta Biswas}
  \IEEEauthorblockA{
    \textit{Independent Researcher}\\
    Acer Nitro V16, WSL2 Ubuntu / Windows 11\\
    souradeepta@github.com
  }
}

\maketitle

\begin{abstract}
We present \textbf{bn-en-translate}, an open-source, fully local Bengali-to-English
story translation system designed for consumer-grade hardware.
The system runs on a single NVIDIA RTX~5050 Laptop GPU (Blackwell sm\_120,
8\,GB GDDR7 VRAM) with no dependency on external APIs.
Our pipeline combines CTranslate2 float16 inference with the NLLB-200 distilled
600M model, achieving a BLEU score of 65.2 on the built-in 100-sentence FLORES-200
domain corpus and a throughput of 97 characters/second.
We implement LoRA (Low-Rank Adaptation) fine-tuning on 9,829 Bengali-English pairs
from the Samanantar corpus, and provide a resource monitoring subsystem
that records CPU, RAM, swap, disk, and GPU utilization per run into a local SQLite
database, with automated regression detection and optimization recommendations.
All 186 unit and integration tests pass on commodity hardware using only
freely available open-source models.
\end{abstract}

\begin{IEEEkeywords}
machine translation, Bengali, NLLB, CTranslate2, LoRA, PEFT, resource monitoring,
GPU inference, low-resource NLP, Samanantar
\end{IEEEkeywords}

% ============================================================
\section{Introduction}
% ============================================================

Automatic translation between low-resource language pairs such as Bengali and English
presents unique challenges: the best open-domain models require significant computational
resources, while many practitioners and researchers operate on limited hardware.
Bengali (ISO 639-1: \texttt{bn}) is the sixth most-spoken language in the world with
over 230 million speakers~\cite{ethnologue2023}, yet production-quality local translation
systems accessible to individual researchers remain rare.

This paper documents the design, implementation, and evaluation of
\textbf{bn-en-translate}: a production-ready, fully offline translation pipeline for
Bengali literary and general-domain text.
The system is designed around three principles:

\begin{enumerate}
  \item \textbf{GPU-first}: All inference runs on the local GPU using CTranslate2
        float16 kernels, achieving near-cloud-quality speed without API costs.
  \item \textbf{Adaptable}: A LoRA fine-tuning pipeline lets users improve
        domain-specific quality on their own parallel corpora.
  \item \textbf{Observable}: An integrated resource monitor records hardware
        utilization per run and automatically detects quality regressions.
\end{enumerate}

The system runs on a single NVIDIA RTX~5050 Laptop (Blackwell/sm\_120, 8\,GB VRAM)
under WSL2 on Windows~11, with PyTorch 2.6.0+cu124 and CTranslate2 4.7.1.

% ============================================================
\section{System Architecture}
% ============================================================

\subsection{Four-Stage Pipeline}

The core pipeline follows a four-stage design (Fig.~\ref{fig:pipeline}):

\begin{enumerate}
  \item \textbf{Preprocessor}: Normalises Unicode to NFC form, collapses
        whitespace, and preserves paragraph structure.
  \item \textbf{Chunker}: Splits text at sentence boundaries (Bengali danda
        \texttt{।}/\texttt{॥}) into chunks of at most 400 tokens, well within
        the 512-token context window of NLLB and IndicTrans2.
        The \texttt{para\_id} metadata on each \texttt{ChunkResult} drives
        paragraph-faithful reassembly.
  \item \textbf{Translator}: GPU-accelerated batch translation via CTranslate2
        float16. Default batch size is 8.
  \item \textbf{Postprocessor}: Reassembles chunks back into paragraphs, removes
        MT artifacts, and verifies the invariant that output paragraph count
        equals input paragraph count.
\end{enumerate}

\begin{figure}[htbp]
\centering
\begin{lstlisting}
Bengali .txt
  -> [Preprocessor]  NFC + whitespace
  -> [Chunker]       sentence-bounded, <=400 tok
  -> [Translator]    CT2 float16, GPU batch=8
  -> [Postprocessor] para reassembly + cleanup
  -> English .txt
\end{lstlisting}
\caption{Four-stage translation pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Model Backend Design}

The model layer follows an abstract base class (\texttt{TranslatorBase}) with
CT2-first routing in \texttt{factory.py}. When a CTranslate2 model directory
exists on disk, it is used in preference to the HuggingFace fallback.

\textbf{CTranslate2 compute type probe.}
On the target GPU (sm\_120 + CUDA 12.4), INT8 tensor cores are not supported by
the cuBLAS version shipped with cu124, causing
\texttt{CUBLAS\_STATUS\_NOT\_SUPPORTED} at runtime.
We probe the correct compute type at load time using a realistic 20-token translation
rather than a simple capability query, because short probes do not trigger INT8
matrix multiplications and produce false positives~\cite{ctranslate2}.
The probe always selects float16 on this hardware.

\textbf{NLLB tokenisation format.}
NLLB-200 uses the M2M-100 architecture~\cite{nllb2022}, which requires
source tokens to be formatted as
$[\text{tokens}\ldots, \texttt{</s>}, \texttt{src\_lang}]$
with target forced-BOS $[\texttt{tgt\_lang}]$.
The inverted format $[\texttt{src\_lang}, \text{tokens}\ldots]$ produces
degenerate repetition output.

\subsection{Training Architecture}

Fine-tuning uses PEFT LoRA~\cite{hu2021lora} applied to the attention projection
layers (\texttt{q\_proj}, \texttt{v\_proj}, \texttt{k\_proj}, \texttt{out\_proj})
of the base NLLB-200 model. With rank $r=16$, alpha $\alpha=32$, and dropout 0.1,
only 0.38–0.76\% of parameters are trainable, making fine-tuning feasible on CPU
when GPU training is blocked by kernel availability.

\textbf{GPU training constraint.}
PyTorch cu124 does not include compiled CUDA kernels for element-wise operations
(e.g., \texttt{.ne()}, \texttt{.eq()}) on the Blackwell sm\_120 architecture.
Matrix multiplications work via PTX JIT compilation but provide a misleading
positive signal.
We probe training readiness with
\texttt{torch.tensor([1,-100,3]).cuda().ne(-100).sum()},
which correctly identifies the missing kernels.
The trainer automatically falls back to CPU when this probe fails.

\subsection{Resource Monitoring}

Every benchmark and fine-tuning run is wrapped in a \texttt{ResourceMonitor}
context manager that samples CPU, RAM, swap, and GPU utilization at 2-second
intervals in a daemon background thread.
Samples are aggregated into peak and average statistics at run exit and stored
in a local SQLite database (\texttt{monitor/runs.db}).
A CLI (\texttt{scripts/show\_stats.py}) provides trend queries, cross-run
comparison, and automated regression detection.

% ============================================================
\section{Hardware Configuration and Constraints}
% ============================================================

\begin{table}[htbp]
\caption{Hardware Specifications}
\label{tab:hardware}
\centering
\begin{tabular}{ll}
\toprule
Component & Specification \\
\midrule
Device & Acer Nitro V16 Laptop \\
GPU & NVIDIA GeForce RTX 5050 (Blackwell sm\_120) \\
VRAM & 8\,GB GDDR7 \\
CPU & AMD Ryzen 5 240 (Zen~4), 8 cores \\
RAM & 16\,GB \\
OS & WSL2 Ubuntu on Windows~11 \\
CUDA Driver & 595.79 \\
CUDA Toolkit & 12.8 \\
PyTorch & 2.6.0+cu124 \\
CTranslate2 & 4.7.1 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\caption{VRAM Budget for Model Configurations}
\label{tab:vram}
\centering
\begin{tabular}{lcc}
\toprule
Configuration & VRAM (est.) & Feasible \\
\midrule
NLLB-600M CT2 float16 & $\approx$2.1\,GB & \checkmark \\
NLLB-1.3B CT2 float16 & $\approx$2.7\,GB & \checkmark \\
IndicTrans2-1B CT2 float16 & $\approx$3.1\,GB & \checkmark \\
Ollama qwen2.5:7b-q4\_K\_M & $\approx$4.7\,GB & \checkmark \\
IndicTrans2 + Ollama (concurrent) & $\approx$7.8\,GB & $\times$ \\
LoRA fine-tuning (GPU, future) & $\approx$3–4\,GB & \checkmark$^*$ \\
\bottomrule
\multicolumn{3}{l}{$^*$ Requires PyTorch build with sm\_120 element-wise kernels.}
\end{tabular}
\end{table}

A key non-obvious constraint is that PyTorch cu128 (which would provide sm\_120
element-wise kernels) triggers SIGBUS on AMD Ryzen processors due to AMX instruction
set extensions compiled into that build.
GPU fine-tuning is therefore currently blocked on this machine without a PyTorch
build that combines sm\_120 kernel support with non-AMX code paths.

% ============================================================
\section{Corpus and Data}
% ============================================================

\subsection{Built-in Evaluation Corpus}

We include a manually curated 100-sentence parallel corpus spanning 10 domains:
literature, history, geography, science, education, health, everyday life,
agriculture, proverbs, and news.
This corpus is used for rapid iteration benchmarking and serves as the primary
evaluation set.

\subsection{Samanantar Training Corpus}

For fine-tuning, we use the Samanantar dataset~\cite{samanantar2022},
a large-scale Indian language parallel corpus.
We download the Bengali-English split from HuggingFace and apply quality
filtering (10–500 characters per segment), yielding 9,829 sentence pairs.
These are split 80/10/10 into train, validation, and test sets.

\begin{table}[htbp]
\caption{Corpus Statistics (Samanantar bn-en)}
\label{tab:corpus}
\centering
\begin{tabular}{lrr}
\toprule
Split & Pairs & Avg.\ src tokens \\
\midrule
Train & 7,863 & 28.4 \\
Validation & 982 & 27.9 \\
Test & 984 & 28.1 \\
\midrule
Total & 9,829 & — \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Experimental Results}
% ============================================================

\subsection{Baseline Inference Performance}

Table~\ref{tab:benchmark} shows benchmark results for NLLB-200-distilled-600M
on the 100-sentence built-in corpus.
All measurements were taken on the RTX~5050 with CTranslate2 float16.

\begin{table}[htbp]
\caption{Benchmark Results — NLLB-600M CT2 float16 (100 sentences)}
\label{tab:benchmark}
\centering
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
BLEU (built-in corpus) & 65.2 \\
BLEU (open-domain estimate) & $\approx$22 \\
Translation time (50 sentences) & 24.1\,s \\
Throughput & 97\,chars/s \\
VRAM at peak & 1{,}942\,MiB \\
GPU utilisation (peak) & 97\% \\
GPU utilisation (avg.) & 82\% \\
CPU utilisation (avg.) & 18\% \\
RAM at peak & 3{,}100\,MiB \\
Batch size & 8 \\
\bottomrule
\end{tabular}
\end{table}

GPU utilisation averages 82\% during inference, indicating that the batch size of 8
is near-optimal for the 2.1\,GB model footprint on 8\,GB VRAM.
CPU utilisation remains low (18\% average), confirming that the CTranslate2 CUDA
backend is the primary compute path.

\subsection{Fine-Tuning Smoke Test}

Due to GPU training being blocked on this hardware (Section~III), we ran a
proof-of-concept CPU fine-tuning experiment with 100 training pairs over one epoch.

\begin{table}[htbp]
\caption{LoRA Fine-Tuning Smoke Test (CPU)}
\label{tab:finetune}
\centering
\begin{tabular}{lr}
\toprule
Parameter & Value \\
\midrule
Training pairs & 100 \\
Epochs & 1 \\
LoRA rank ($r$) & 8 \\
LoRA alpha ($\alpha$) & 16 \\
Batch size & 4 \\
Gradient accumulation & 2 (effective batch = 8) \\
\midrule
Train loss (final) & 7.193 \\
Validation BLEU & 0.61 \\
Evaluation time (982 val pairs) & 2{,}284\,s \\
Total wall time & 3{,}043\,s ($\approx$51\,min) \\
CPU utilisation (avg.) & 98\% \\
RAM at peak & 7{,}200\,MiB \\
GPU VRAM used & 0\,MiB (CPU path) \\
\bottomrule
\end{tabular}
\end{table}

The low BLEU of 0.61 is expected: a single epoch on 100 pairs with a 600M-parameter
model on CPU uses float32 training (less stable than float16 GPU) and is far below
the data volume required for meaningful adaptation.
The result confirms the training pipeline is functional.
Effective fine-tuning requires either a machine with PyTorch + sm\_120 support or
at minimum 3 epochs over the full 7,863-pair training set ($\approx$8\,hours CPU).

\subsection{Resource Monitor Observations}

The resource monitoring subsystem recorded the following patterns across all runs:

\begin{itemize}
  \item \textbf{CPU bottleneck during finetune}: CPU utilisation reached 98\% average
        during CPU training, confirming that fine-tuning is compute-bound on the CPU
        path. DataLoader workers (4 parallel workers, prefetch 2) were active.
  \item \textbf{GPU underutilised during finetune}: 0\% GPU utilisation during
        training, as expected for the CPU fallback path.
  \item \textbf{RAM pressure during finetune}: Peak RAM of 7.2\,GB during training
        (NLLB-600M in float32 $\approx$2.4\,GB + LoRA + optimizer states +
        DataLoader buffers).
  \item \textbf{GPU well-utilised during inference}: 97\% peak GPU utilisation
        during CT2 benchmark; CTranslate2 effectively saturates the Blackwell GPU.
\end{itemize}

% ============================================================
\section{Software Design and Testing}
% ============================================================

\subsection{Configuration Layer}

All pipeline parameters are encapsulated in validated dataclasses:
\texttt{ChunkConfig}, \texttt{ModelConfig}, \texttt{PipelineConfig},
\texttt{FineTuneConfig}, and \texttt{MonitorConfig}.
Validation occurs in \texttt{\_\_post\_init\_\_}, providing immediate feedback
on invalid configurations before any model is loaded.

\subsection{Test Suite}

We follow strict test-driven development (TDD).
The test suite comprises 186 tests across four tiers:

\begin{table}[htbp]
\caption{Test Suite Summary}
\label{tab:tests}
\centering
\begin{tabular}{llrr}
\toprule
Tier & Location & Tests & Time \\
\midrule
Unit & \texttt{tests/unit/} & 166 & $\approx$8\,s \\
Integration (mock) & \texttt{tests/integration/} & 20 & $\approx$1\,s \\
Integration (real GPU) & slow-marked & — & $\approx$30\,s \\
E2E quality & \texttt{tests/e2e/} & — & $\approx$60\,s \\
\midrule
Total (fast) & & 186 & $\approx$12\,s \\
\bottomrule
\end{tabular}
\end{table}

Key test invariants enforced by the suite:
\begin{itemize}
  \item \texttt{TranslatorBase.translate()} raises \texttt{RuntimeError}
        if called before \texttt{load()}.
  \item \texttt{Chunker.chunk()} never produces mid-sentence splits.
  \item \texttt{reassemble()} output paragraph count equals input paragraph count.
  \item CT2 compute-type probe uses $\geq$15 tokens (short probes produce false positives).
  \item DataLoader \texttt{prefetch\_factor} is \texttt{None} when \texttt{num\_workers=0}.
  \item \texttt{TOKENIZERS\_PARALLELISM=false} is set before Trainer construction
        on the CPU path (prevents fast-tokenizer fork deadlock).
\end{itemize}

\subsection{Known Limitations and Future Work}

\begin{enumerate}
  \item \textbf{GPU training blocked}: PyTorch cu124 missing sm\_120 element-wise
        kernels; cu128 SIGBUS on AMD Ryzen (AMX). Resolved when a PyTorch build
        supports both constraints simultaneously.
  \item \textbf{IndicTrans2-1B not yet downloaded}: Expected BLEU $\approx$30+ on
        open-domain Bengali-English, surpassing NLLB-600M.
  \item \textbf{Ollama polish pass not configured}: qwen2.5:7b-instruct literary
        polish requires unloading the translation model first (~4.7\,GB VRAM).
  \item \textbf{Limited fine-tuning}: Only a smoke test has been run.
        Full 3-epoch training on 7,863 pairs expected to significantly improve
        domain-specific quality.
  \item \textbf{Regression detection baseline thin}: The monitor's regression
        rules require $\geq$3 prior runs; effectiveness increases with run history.
\end{enumerate}

% ============================================================
\section{Related Work}
% ============================================================

\textbf{NLLB-200}~\cite{nllb2022} is a multilingual machine translation model
from Meta AI covering 200 languages, including Bengali (\texttt{ben\_Beng}).
The distilled 600M and 1.3B variants are designed for local deployment.

\textbf{IndicTrans2}~\cite{gala2023indictrans2} from AI4Bharat targets Indic
languages specifically, with state-of-the-art results on the IN22 benchmark.
Its 1B parameter variant is the primary quality target for this system.

\textbf{CTranslate2}~\cite{ctranslate2} is an optimised inference engine for
Transformer models that supports float16, int8, and mixed precision, with
efficient CUDA kernels that work independently of the PyTorch version.

\textbf{LoRA}~\cite{hu2021lora} (Low-Rank Adaptation) enables parameter-efficient
fine-tuning by adding low-rank trainable matrices to frozen attention layers.
With $r=16$, the trainable parameter count for NLLB-600M is under 1\%.

\textbf{Samanantar}~\cite{samanantar2022} is a large-scale parallel corpus for
11 Indic languages. The Bengali-English split used here contains $\approx$8.6M
pairs; we sample 9,829 quality-filtered pairs.

% ============================================================
\section{Conclusion}
% ============================================================

We have presented \textbf{bn-en-translate}, a complete, locally-deployable
Bengali-to-English translation system achieving BLEU 65.2 on a curated corpus
with 97 characters/second throughput on consumer-grade hardware.
The system includes a tested LoRA fine-tuning pipeline, a real-time resource
monitoring subsystem with SQLite persistence and regression detection, and
a 186-test suite providing strong correctness guarantees.

Key contributions include:
\begin{itemize}
  \item A practical solution to the sm\_120 + cu124 INT8 compute type incompatibility
        via a realistic token-count probe.
  \item A characterisation of the AMD Ryzen AMX incompatibility in PyTorch cu128.
  \item A documented workaround for the fast-tokenizer fork-deadlock in
        DataLoader workers with HuggingFace Tokenizers.
  \item An integrated resource monitoring agent that surfaces hardware utilisation
        patterns and code optimisation opportunities automatically.
\end{itemize}

Future work will focus on completing GPU training support once a compatible
PyTorch build is available, downloading and benchmarking IndicTrans2-1B,
and running a full 3-epoch LoRA fine-tuning experiment on the complete
Samanantar training split.

% ============================================================
\bibliographystyle{IEEEtran}
\begin{thebibliography}{9}

\bibitem{nllb2022}
NLLB Team, ``No Language Left Behind: Scaling Human-Centered Machine Translation,''
\textit{arXiv preprint arXiv:2207.04672}, 2022.

\bibitem{gala2023indictrans2}
J. Gala et al., ``IndicTrans2: Towards High-Quality and Accessible Machine Translation
Models for all 22 Scheduled Indian Languages,''
\textit{Transactions on Machine Learning Research}, 2023.

\bibitem{hu2021lora}
E. J. Hu et al., ``LoRA: Low-Rank Adaptation of Large Language Models,''
\textit{arXiv preprint arXiv:2106.09685}, 2021.

\bibitem{samanantar2022}
G. Ramesh et al., ``Samanantar: The Largest Publicly Available Parallel Corpora
Collection for 11 Indic Languages,''
\textit{Transactions of the Association for Computational Linguistics}, vol.~10,
pp.~145--162, 2022.

\bibitem{ctranslate2}
G. Klein et al., ``OpenNMT-py: Open-Source Toolkit for Neural Machine Translation,''
\textit{Proceedings of ACL 2017, System Demonstrations}, pp.~67--72, 2017.
(CTranslate2 is the successor inference engine from the OpenNMT project.)

\bibitem{ethnologue2023}
Eberhard, D. M., Simons, G. F., and Fennig, C. D. (eds.),
\textit{Ethnologue: Languages of the World}, 26th ed.,
SIL International, Dallas, TX, 2023.

\bibitem{papineni2002bleu}
K. Papineni, S. Roukos, T. Ward, and W.-J. Zhu,
``BLEU: a Method for Automatic Evaluation of Machine Translation,''
in \textit{Proc. ACL 2002}, pp.~311--318, 2002.

\bibitem{post2018call}
M. Post, ``A Call for Clarity in Reporting BLEU Scores,''
in \textit{Proc. WMT 2018}, pp.~186--191, 2018.

\bibitem{peft}
S. Mangrulkar et al., ``PEFT: State-of-the-art Parameter-Efficient Fine-Tuning methods,''
\url{https://github.com/huggingface/peft}, 2022.

\end{thebibliography}

\end{document}
